SIMD\slash SIMT-vectorized Sparse Chemical Kinetic Jacobian\slash Thermo-chemical Source Term Evaluation

### 2018-03-09 paper.tex
s/runtime/run time for describing execution times vs e.g., the Intel runtime"

%====================================================================
\begin{abstract} % not to exceed 200 words
A code generation platform for single-instruction, multiple-data (SIMD) and single-instruction, multiple thread (SIMT) vectorized thermo-chemical source term and sparse chemical kinetic Jacobian evaluation was developed and validated for a shallow range of chemical kinetic models.
\todo[inline]{this}
\end{abstract}

@@

As the importance of detailed chemical kinetics for predictive reactive-flow simulations has become recognized~\cite{LU2009192}, the size and complexity of chemical kinetic models describing current and next-generation fuels relevant to transportation and power generation have increased greatly; e.g., a recent biodiesel model~\cite{WESTBROOK2011742} consists of \textasciitilde\num{3500} chemical species, and over \num{17000} reactions.
Moreover, the cost of evaluating the chemical source terms scales linearly with the size of the model, while evaluation and factorization of the chemical kinetic Jacobian---if naively implemented---scale quadratically and cubically with the number of species in the model~\cite{LU2009192}.
These factors often make the use of detailed chemical kinetics prohibitive in practice; e.g., in a recent direct numerical simulation~\cite{Spafford:2010aa} utilizing a 22-species model, evaluation of reaction rates consumed around half of the total runtime.
In addition, chemical kinetic Jacobian evaluation and factorization---required for most common implicit integration techniques utilized to deal with numerical stiffness---is a bottleneck for even modestly-sized chemical models in realistic reactive flow simulations, and necessitates use of other cost-reduction strategies~\cite{LU2009192}.

@@

\subsection{Performance}
\label{S:results}
The performance studies in this work were run on the platforms listed in~\cref{t:cpus,t:gpus}.
Runtimes in each case were averaged over ten runs, each using the same set of PaSR conditions utilized in validation.
The OpenMP Jacobian\slash source-rate kernels, as well as the OpenMP\slash OpenCL wrapping code---responsible for initializing\slash transferring memory, reading input, etc.---was compiled with \texttt{gcc v5.4.0} on the \avx/\slash\gpunew/ platforms, and \texttt{gcc v4.8.5} on the \sse/\slash\gpuold/ machines.
The optimization level ``\texttt{-O3 -mtune=native}'' was used, and no ``fast-math'' OpenCL optimizations were enabled.
Additionally, the exact form of the Jacobian (as opposed the the ``approximate'' form discussed in~\cref{S:sparsity}) was used in all cases.
Finally, unless stated otherwise: the performance results utilized the \conp/ assumption, a vector width of \num{8}\slash\num{128} and ``C''\slash ``F''-ordered data for the CPU\slash GPU cases respectively; the runtimes reported are for the number of conditions specified in~\cref{T:error} and include data transfer overhead to\slash from internal buffers used in \texttt{pyJac}.
The effects of choice of vector-width, data-ordering and differences between \conp/ and \conv/ evaluations on the CPU\slash GPU will be explored in~\cref{S:source_results}.

@@

In~\cref{F:intel_source_nonorm}, the mean runtime per initial condition is shown for both the~\avx/ and~\sse/ CPUs, using both Intel OpenCL and OpenMP.
This normalization of the runtime by the number of initial conditions tested is chosen to account for the varying numbers of conditions in the PaSR databases for each model (\cref{t:models}).
For both CPUs, the OpenMP implementation is the slowest over all models; interestingly the unvectorized (strictly parallely) Intel OpenCL code is slightly faster than OpenMP in all cases.
As expected, the \avx/ machine is faster than the \sse/ CPU for the strictly parallel cases, \SIrange{1.82}{2.13}{$\times$} for the unvectorized OpenCL case, and \SIrange{1.72}{1.85}{$\times$} for OpenMP.
Additionally, the shallow-vectorized OpenCL code is significantly faster than either the OpenMP or unvectorized OpenCL codes on both processors.

\begin{figure}[htbp]
   \centering
  \begin{subfigure}[t]{0.48\linewidth}
      \includegraphics[width=\textwidth]{intel_source_nonorm.pdf}
      \caption{The mean runtime per condition for each chemical model using the Intel OpenCL and OpenMP on both CPUs studied.}
      \label{F:intel_source_nonorm}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.48\linewidth}
      \includegraphics[width=\textwidth]{intel_source.pdf}
      \caption{The speedup achieved over the baseline OpenMP parallelization by both the unvectorized and shallow-vectorized Intel OpenCL codes; the speedup is presented on per-machine basis.}
      \label{F:intel_source}
  \end{subfigure}
  \\
  \begin{subfigure}[t]{0.48\linewidth}
      \includegraphics[width=\textwidth]{pocl_source_nonorm.pdf}
      \caption{The mean runtime per condition of the Portable OpenCL runtime compared to OpenMP parallelization.}
      \label{F:pocl_source}
  \end{subfigure}
 \caption{Mean runtimes per condition and speedups achieved by the various CPU OpenCL runtimes compared to OpenMP parallelization for each chemical model studied. The names in the legends correspond to the identifiers listed in~\cref{t:cpus}.}
 \label{F:cpu_source}
\end{figure}

In~\cref{F:intel_source}, the extent of this speedup is detailed; we note that the speedup displayed is calculated per-machine, e.g., the \avx/ shallow-vectorized code speedup is relative to OpenMP on the same CPU such that the speedup of the shallow-vectorized OpenCL code can be readily assessed.
On both machines, the unvectorized OpenCL code is faster than the baseline parallel OpenMP code; by \SIrange{1.30}{1.51}{$\times$} on the \avx/ CPU, and \SIrange{1.25}{1.27}{$\times$} on the \sse/ machine.
Additionally, the shallow-vectorized OpenCL code is \textasciitilde\SIrange{2.53}{2.92}{$\times$} and \SIrange{3.40}{4.08}{$\times$} faster than the OpenMP code for the \sse/ and \avx/ machines respectively.
In contrast,~\cref{F:pocl_source} shows the mean runtime per condition of deep, shallow and unvectorized OpenCL codes on the POCL runtime, as compared to OpenMP parallelization.
No speedup is achieved for either vectorization type on either CPU; indeed the OpenMP case is faster in both cases.
The reasons for this lack of vectorization are quite technical and somewhat outside the scope of this work; however, briefly speaking vectorization might be achieved on POCL via a combination of fixed-array strides---rather than (potentially) varying with the number of initial conditions being evaluated, as currently is implemented---use of relaxed math (``\texttt{-cl-fast-relaxed-math}''), explicit OpenCL vector types and newer versions of LLVM\slash POCL (specifically, potential improvements to the vectorization module).
This will be explored more in future work, but for the moment POCL is still quite useful as a validation tool.

\Cref{F:gpu_source} shows the performance of the source term evaluation on the GPUs listed in~\cref{t:gpus}.
In~\cref{F:gpu_source_scaling}, the number of initial conditions over which the source terms are evaluated on the \gpunew/ GPU is varied in order to determine the effect on the mean runtime per condition; the mean runtime continues to decrease until \textasciitilde\num{e4} conditions for all chemical models (except the \ce{H2}\slash\ce{CO} model, which is closer to \textasciitilde\num{3e4} conditions), at which point the GPU becomes saturated and performance levels off.
Interestingly, the last data-point for the GRI-Mech 3.0 shows an increase in runtime per condition, this is likely due to multiple data transfers to\slash from the GPU required to complete the source-rates in this case; similar (if smaller) increases in runtime per condition can be seen for the \ce{iC5H11OH} model near \num{e4} and \num{e5} conditions.\todo{rerunning}
\Cref{F:gpu_source_speedup} shows the speedup in source term evaluations the \gpunew/ GPU over the \gpuold/ GPU for the maximum number of conditions in~\cref{F:gpu_source_scaling} for two different vector widths (a.k.a., the block size in NVIDIA terminology), \num{64} and \num{128}; the fastest \gpunew/ case (a vector width of 128) is \SIrange{1.34}{4.89}{$\times$} faster than the slowest case (either \gpunew/ or \gpuold/ with a vector width of \num{64}) depending on the chemical model in question.
Additionally,~\cref{F:gpu_source_speedup} shows that varying the vector width has minimal effect on performance for most \gpuold/ cases (excepting \ce{H2}\slash\ce{CO}), but a large effect on the performance of the \gpunew/ GPU; this is likely caused by higher occupancy~\cite{occupancy} on the \gpunew/ GPU, however it is unclear exactly how NVIDIA's OpenCL runtime balances the registers\slash warps per streaming-multiprocessor, as controls occupancy in CUDA~\cite{occupancy}; this is a topic that merits future investigation.
\begin{figure}[htbp]
   \centering
  \begin{subfigure}[t]{0.48\linewidth}
      \includegraphics[width=\textwidth]{gpu_source_scaling.pdf}
      \caption{The mean runtime per condition for each chemical model on the \gpunew/ GPU as a function of the number of initial conditions tested.}
      \label{F:gpu_source_scaling}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.48\linewidth}
      \includegraphics[width=\textwidth]{gpu_source_speedup.pdf}
      \caption{The speedup achieved on the \gpunew/ versus the \gpuold/ GPU; the names correspond to the identifiers listed in~\cref{t:gpus}}
      \label{F:gpu_source_speedup}
  \end{subfigure}
  \caption{\texttt{pyJac} source term evaluation performance on the NVIDIA GPUs}
  \label{F:gpu_source}
\end{figure}

\Cref{f:source_permutate} shows the effect of changing data-ordering patterns, the \conp/ or \conv/ formulation, and the CPU vector-width size on the performance of source term evaluation in \texttt{pyJac}.
In~\cref{F:source_conpvsconv}, we see that choice of the constant-pressure or constant-volume formulation has little to no effect on runtime for OpenMP as well as the shallow-vectorized\slash unvectorized Intel OpenCL codes on the \avx/ machine.
Generally speaking, the difference between the \conp/ and \conv/ formulations has only a marginal effect on performance regardless of CPU\slash GPU choice.

\begin{figure}[htbp]
   \centering
  \begin{subfigure}[t]{0.48\linewidth}
      \includegraphics[width=\textwidth]{source_conpvsconv.pdf}
      \caption{The mean runtime per condition for each chemical model using both the \conp/ and \conv/ formulations on Intel OpenCL and OpenMP on the \avx/ CPU.}
      \label{F:source_conpvsconv}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.48\linewidth}
      \includegraphics[width=\textwidth]{source_cvsf.pdf}
      \caption{The speedup achieved by ``C'' ordering over ``F'' ordering for Intel OpenCL and OpenMP on the \avx/ CPU.  The speedup presented is calculated per-language (OpenMP and OpenCL) to better assess the effect of the data-ordering.}
      \label{F:source_cvsf}
  \end{subfigure}
  \\
  \begin{subfigure}[t]{0.48\linewidth}
      \includegraphics[width=\textwidth]{source_gpu_cvsf.pdf}
      \caption{The affect on performance of ``C'' vs ``F''-ordering for the shallow-vectorized NVIDIA OpenCL code on both GPUs with a vector width of 128.  The speedup presented is calculated per-GPU to better assess the effect of the data-ordering on each GPU.}
      \label{F:source_gpu_cvsf}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.48\linewidth}
      \includegraphics[width=\textwidth]{source_vector_width.pdf}
      \caption{The effect of vector-width on ``C''-ordered shallow-vectorized Intel OpenCL source term evaluations in the \avx/ CPU.}
      \label{F:source_vector_width}
  \end{subfigure}
  \caption{The effect of the \conp/ and \conv/ formulations, ``C'' and ``F'' data-ordering, and CPU vector-width on source term evaluation performance in \texttt{pyJac}.
	   It is noted the shallow-vectorized ``C'' ordered OpenCL cases corresponds to the vectorized data ordering described in~\cref{S:data}}
  \label{f:source_permutate}
\end{figure}

@@

\Cref{F:source_gpu_cvsf} shows the performance effect of the ``C'' versus ``F''-ordering on source term evaluation performance on both GPUs, with the speedup presented per-GPU.
\todo[inline]{rewrite this once I get the re-run performance data for the source terms}
In general, the ``C'' and ``F''-ordered shallow-vectorizations are nearly equivalent on both GPUs while the ``C'' and ``F''-ordered data are nearly equivalent for the \gpuold/ GPU, with the exception of the isopentanol model where the ``C''-ordered data is up to \SI{1.1}{$\times$} faster on the \gpunew/ (we note that there was high variablity for the \gpuold/ in this case). \todo{again, pending rerun on \gpuold/}
The rough equivalence between the ``C'' and ``F''-ordered codes GPUs is counter to what one might expect, as typically speaking coalesced memory access in a shallow-vectorization is easier to achieve if ``F''-ordering is used (see~\cref{S:data}).
However, the vectorized-data ordering here ensures that not only are memory accesses aligned to the vector width size and thus, likely coalesced. % but the improved data-locality, coupled with the larger cache sizes of the \gpunew/ GPU result in this modest speedup for ``C''-data ordering.

@@

\begin{figure}[htbp]
   \centering
  \begin{subfigure}[t]{0.48\linewidth}
      \includegraphics[width=\textwidth]{source_parallel_scaling.pdf}
      \caption{The strong parallel scaling efficiency (as defined in~\cref{e:strong_scaling}) of source term evaluation for Intel OpenCL\slash OpenMP on the \avx/ machine.}
      \label{F:source_parallel_scaling}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.48\linewidth}
      \includegraphics[width=\textwidth]{source_crossover.pdf}
      \caption{The mean runtime per-condition of OpenMP source term evaluation on different cores compared to a shallow-vectorized Intel OpenCL evaluation on \num{1} (solid lines) and \num{4} (dashed lines) cores on \avx/ machine.}
      \label{F:source_crossover}
  \end{subfigure}
  \\
  \begin{subfigure}[t]{0.48\linewidth}
      \includegraphics[width=\textwidth]{source_simd_efficiency.pdf}
      \caption{The SIMD efficiency (\cref{e:simd_efficiency}) of source term evaluation for the Intel OpenCL runtime on a single core of the \avx/ CPU.}
      \label{F:source_simd_scaling}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.48\linewidth}
      \includegraphics[width=\textwidth]{source_sse_simd_efficiency.pdf}
      \caption{The SIMD efficiency (\cref{e:simd_efficiency}) of source term evaluation for the Intel OpenCL runtime on a single core of the \sse/ CPU.}
      \label{F:source_sse_simd_scaling}
  \end{subfigure}
  \caption{The parallel scaling efficiency and SIMD efficiency of source term evaluation for Intel OpenCL on the \avx/ and \sse/ CPUs.}
  \label{F:source_scaling}
\end{figure}

Finally,~\cref{F:source_scaling} displays the (strong) parallel scaling efficiency and SIMD efficiency for the CPU platforms.
The strong parallel scaling efficiency is defined as:
\begin{equation}
 \label{e:strong_scaling}
 \varepsilon = \frac{\bar{t}_{1}}{N \bar{t}_{N}}
\end{equation}
where $\bar{t}_{N}$ is the mean runtime per condition on $N$ CPU cores, and $\bar{t}_{1}$ the same on a single CPU core.
The strong parallel scaling efficiency measures the speedup due to the use of additional CPU cores, as a fraction of linear speedup; strong scaling tends to decrease with the number of processors used due to memory bandwidth limitations and decreasing computation work allocated per CPU core~\cite{strong_scaling}.

\Cref{F:source_parallel_scaling} shows the strong parallel scaling efficiency of source-rate in \texttt{pyJac} on the \avx/ machine for both the shallow-vectorized Intel OpenCL and OpenMP codes.
In general, the \ce{H2}\slash\ce{CO} mechanism has the worst scaling efficiency for Intel OpenCL; although the mechanism scalings vary somewhat for OpenMP, there is no clear trend.
It is likely the poor scaling of the \ce{H2}\slash\ce{CO} for Intel OpenCL results from both its relatively small size, and few falloff\slash chemically activated reaction (in particular, the additional expensive logarithm and exponential evaluations that accompany them); i.e., there is simply not enough work to fully utilize the SIMD intructions.
Additionally, OpenMP tends to have higher scalings then the shallow-vectorized OpenCL code, e.g., \textasciitilde\num{0.9} and \textasciitilde\num{0.75} for \num{4} and \num{8} CPU cores respectively, compared to just \numrange{0.66}{0.72} and \numrange{0.44}{0.48} for OpenCL; although not pictured (to keep the figure readable), the unvectorized Intel OpenCL code has only slightly worse scaling than the OpenMP code, hence the poorer scaling is unique to the shallow-vectorizated code.
This is due to better performance of the shallow-vectorized OpenCL code, coupled with the relatively small amounts of work associated with source-rate.
To illustrate this,~\cref{F:source_crossover} shows the mean runtime per-condition of the OpenMP source term evaluations for \numrange{2}{16} cores, compared to the shallow-vectorized OpenCL code on \num{1} (solid lines) and \num{4} (dashed lines) four cores on the \avx/ machine.
For all chemical models, the mean runtime per-condition (and hence the computational work allocated per-core, one of the key-drivers of parallel scaling efficiency~\cite{strong_scaling}) of OpenMP running on \num{4} cores is roughly equal to that of the shallow-vectorized OpenCL code on a single core.
Similarly, OpenMP running on \num{16} cores is roughly equivalent to the OpenCL code on \num{4} cores.
Therefore, a more fair comparison of parallel scaling efficiencies is to compare OpenCL running on \num{4} cores, to OpenMP on \num{16}; the OpenMP code's parallel efficiency drops to \textasciitilde\num{0.64} for \num{16} cores, similar to OpenCL's parallel scaling efficiency of \numrange{0.66}{0.72} at \num{4} cores.
Indeed, as will be investigated in~\cref{S:jacobian_results} the strong parallel scaling efficiency for sparse Jacobian evaluation---the most computationally intensive task in this work---are far more similar between Intel OpenCL and OpenMP.

>>>>

%====================================================================
\begin{abstract} % not to exceed 200 words
A code generation platform for single-instruction, multiple-data (SIMD) and single-instruction, multiple thread (SIMT) vectorized thermo-chemical source term and sparse\slash dense chemical kinetic Jacobian evaluation was developed and validated for a wide range of chemical kinetic models.

\todo[inline]{this}
\end{abstract}

@@

As the importance of detailed chemical kinetics for predictive reactive-flow simulations has become recognized~\cite{LU2009192}, the size and complexity of chemical kinetic models describing current and next-generation fuels relevant to transportation and power generation have increased greatly; e.g., a recent biodiesel model~\cite{WESTBROOK2011742} consists of \textasciitilde\num{3500} chemical species, and over \num{17000} reactions.
Moreover, the cost of evaluating the chemical source terms scales linearly with the size of the model, while evaluation and factorization of the chemical kinetic Jacobian---if naively implemented---scale quadratically and cubically with the number of species in the model~\cite{LU2009192}.
These factors often make the use of detailed chemical kinetics prohibitive in practice; e.g., in a recent direct numerical simulation~\cite{Spafford:2010aa} utilizing a 22-species model, evaluation of reaction rates consumed around half of the total run time.
In addition, chemical kinetic Jacobian evaluation and factorization---required for most common implicit integration techniques utilized to deal with numerical stiffness---is a bottleneck for even modestly-sized chemical models in realistic reactive flow simulations, necessitating use of other cost-reduction strategies~\cite{LU2009192}.

@@

\subsection{Performance}
\label{S:results}
The performance studies in this work were run on the platforms listed in~\cref{t:cpus,t:gpus}.
Runtimes in each case were averaged over ten runs, each using the same set of PaSR conditions utilized in validation.
The OpenMP Jacobian\slash source-rate kernels, as well as the OpenMP\slash OpenCL wrapping code---responsible for initializing\slash transferring memory, reading input, etc.---was compiled with \texttt{gcc v5.4.0} on the \avx/\slash\gpunew/ platforms, and \texttt{gcc v4.8.5} on the \sse/\slash\gpuold/ machines.
The optimization level ``\texttt{-O3 -mtune=native}'' was used, and no ``fast-math'' OpenCL optimizations were enabled.
Additionally, the exact form of the Jacobian (as opposed the the ``approximate'' form discussed in~\cref{S:sparsity}) was used in all cases.
Finally, unless stated otherwise: the performance results utilized the \conp/ assumption, a vector width of \num{8}\slash\num{128} and ``C''\slash ``F''-ordered data for the CPU\slash GPU cases respectively; the run times reported are for the number of conditions specified in~\cref{T:error} and include data transfer overhead to\slash from internal buffers used in \texttt{pyJac}.
The effects of choice of vector-width, data-ordering and differences between \conp/ and \conv/ evaluations on the CPU\slash GPU will be explored in~\cref{S:source_results}.

@@

In~\cref{F:intel_source_nonorm}, the mean run time per initial condition is shown for both the~\avx/ and~\sse/ CPUs, using both Intel OpenCL and OpenMP.
This normalization of the run time by the number of initial conditions tested is chosen to account for the varying numbers of conditions in the PaSR databases for each model (\cref{t:models}).
For both CPUs, the OpenMP implementation is the slowest over all models; interestingly the unvectorized (strictly parallely) Intel OpenCL code is slightly faster than OpenMP in all cases.
As expected, the \avx/ machine is faster than the \sse/ CPU for the strictly parallel cases, \SIrange{1.82}{2.13}{$\times$} for the unvectorized OpenCL case, and \SIrange{1.72}{1.85}{$\times$} for OpenMP.
Additionally, the shallow-vectorized OpenCL code is significantly faster than either the OpenMP or unvectorized OpenCL codes on both processors.

\begin{figure}[htbp]
   \centering
  \begin{subfigure}[t]{0.48\linewidth}
      \includegraphics[width=\textwidth]{intel_source_nonorm.pdf}
      \caption{The mean run time per condition for each chemical model using the Intel OpenCL and OpenMP on both CPUs studied.}
      \label{F:intel_source_nonorm}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.48\linewidth}
      \includegraphics[width=\textwidth]{intel_source.pdf}
      \caption{The speedup achieved over the baseline OpenMP parallelization by both the unvectorized and shallow-vectorized Intel OpenCL codes; the speedup is presented on per-machine basis.}
      \label{F:intel_source}
  \end{subfigure}
  \\
  \begin{subfigure}[t]{0.48\linewidth}
      \includegraphics[width=\textwidth]{pocl_source_nonorm.pdf}
      \caption{The mean run time per condition of the Portable OpenCL runtime compared to OpenMP parallelization.}
      \label{F:pocl_source}
  \end{subfigure}
 \caption{Mean run times per condition and speedups achieved by the various CPU OpenCL runtimes compared to OpenMP parallelization for each chemical model studied. The names in the legends correspond to the identifiers listed in~\cref{t:cpus}.}
 \label{F:cpu_source}
\end{figure}

In~\cref{F:intel_source}, the extent of this speedup is detailed; we note that the speedup displayed is calculated per-machine, e.g., the \avx/ shallow-vectorized code speedup is relative to OpenMP on the same CPU such that the speedup of the shallow-vectorized OpenCL code can be readily assessed.
On both machines, the unvectorized OpenCL code is faster than the baseline parallel OpenMP code; by \SIrange{1.30}{1.51}{$\times$} on the \avx/ CPU, and \SIrange{1.25}{1.27}{$\times$} on the \sse/ machine.
Additionally, the shallow-vectorized OpenCL code is \textasciitilde\SIrange{2.53}{2.92}{$\times$} and \SIrange{3.40}{4.08}{$\times$} faster than the OpenMP code for the \sse/ and \avx/ machines respectively.
In contrast,~\cref{F:pocl_source} shows the mean run time per condition of deep, shallow and unvectorized OpenCL codes on the POCL runtime, as compared to OpenMP parallelization.
No speedup is achieved for either vectorization type on either CPU; indeed the OpenMP case is faster in both cases.
The reasons for this lack of vectorization are quite technical and somewhat outside the scope of this work; however, briefly speaking vectorization might be achieved on POCL via a combination of fixed-array strides---rather than (potentially) varying with the number of initial conditions being evaluated, as currently is implemented---use of relaxed math (``\texttt{-cl-fast-relaxed-math}''), explicit OpenCL vector types and newer versions of LLVM\slash POCL (specifically, potential improvements to the vectorization module).
This will be explored more in future work, but for the moment POCL is still quite useful as a validation tool.

\Cref{F:gpu_source} shows the performance of the source term evaluation on the GPUs listed in~\cref{t:gpus}.
In~\cref{F:gpu_source_scaling}, the number of initial conditions over which the source terms are evaluated on the \gpunew/ GPU is varied in order to determine the effect on the mean run time per condition; the mean run time continues to decrease until \textasciitilde\num{e4} conditions for all chemical models (except the \ce{H2}\slash\ce{CO} model, which is closer to \textasciitilde\num{3e4} conditions), at which point the GPU becomes saturated and performance levels off.
Interestingly, the last data-point for the GRI-Mech 3.0 shows an increase in run time per condition, this is likely due to multiple data transfers to\slash from the GPU required to complete the source-rates in this case; similar (if smaller) increases in run time per condition can be seen for the \ce{iC5H11OH} model near \num{e4} and \num{e5} conditions.\todo{rerunning}
\Cref{F:gpu_source_speedup} shows the speedup in source term evaluations the \gpunew/ GPU over the \gpuold/ GPU for the maximum number of conditions in~\cref{F:gpu_source_scaling} for two different vector widths (a.k.a., the block size in NVIDIA terminology), \num{64} and \num{128}; the fastest \gpunew/ case (a vector width of 128) is \SIrange{1.34}{4.89}{$\times$} faster than the slowest case (either \gpunew/ or \gpuold/ with a vector width of \num{64}) depending on the chemical model in question.
Additionally,~\cref{F:gpu_source_speedup} shows that varying the vector width has minimal effect on performance for most \gpuold/ cases (excepting \ce{H2}\slash\ce{CO}), but a large effect on the performance of the \gpunew/ GPU; this is likely caused by higher occupancy~\cite{occupancy} on the \gpunew/ GPU, however it is unclear exactly how NVIDIA's OpenCL runtime balances the registers\slash warps per streaming-multiprocessor, as controls occupancy in CUDA~\cite{occupancy}; this is a topic that merits future investigation.
\begin{figure}[htbp]
   \centering
  \begin{subfigure}[t]{0.48\linewidth}
      \includegraphics[width=\textwidth]{gpu_source_scaling.pdf}
      \caption{The mean run time per condition for each chemical model on the \gpunew/ GPU as a function of the number of initial conditions tested.}
      \label{F:gpu_source_scaling}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.48\linewidth}
      \includegraphics[width=\textwidth]{gpu_source_speedup.pdf}
      \caption{The speedup achieved on the \gpunew/ versus the \gpuold/ GPU; the names correspond to the identifiers listed in~\cref{t:gpus}}
      \label{F:gpu_source_speedup}
  \end{subfigure}
  \caption{\texttt{pyJac} source term evaluation performance on the NVIDIA GPUs}
  \label{F:gpu_source}
\end{figure}

\Cref{f:source_permutate} shows the effect of changing data-ordering patterns, the \conp/ or \conv/ formulation, and the CPU vector-width size on the performance of source term evaluation in \texttt{pyJac}.
In~\cref{F:source_conpvsconv}, we see that choice of the constant-pressure or constant-volume formulation has little to no effect on run time for OpenMP as well as the shallow-vectorized\slash unvectorized Intel OpenCL codes on the \avx/ machine.
Generally speaking, the difference between the \conp/ and \conv/ formulations has only a marginal effect on performance regardless of CPU\slash GPU choice.

\begin{figure}[htbp]
   \centering
  \begin{subfigure}[t]{0.48\linewidth}
      \includegraphics[width=\textwidth]{source_conpvsconv.pdf}
      \caption{The mean run time per condition for each chemical model using both the \conp/ and \conv/ formulations on Intel OpenCL and OpenMP on the \avx/ CPU.}
      \label{F:source_conpvsconv}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.48\linewidth}
      \includegraphics[width=\textwidth]{source_cvsf.pdf}
      \caption{The speedup achieved by ``C'' ordering over ``F'' ordering for Intel OpenCL and OpenMP on the \avx/ CPU.  The speedup presented is calculated per-language (OpenMP and OpenCL) to better assess the effect of the data-ordering.}
      \label{F:source_cvsf}
  \end{subfigure}
  \\
  \begin{subfigure}[t]{0.48\linewidth}
      \includegraphics[width=\textwidth]{source_gpu_cvsf.pdf}
      \caption{The affect on performance of ``C'' vs ``F''-ordering for the shallow-vectorized NVIDIA OpenCL code on both GPUs with a vector width of 128.  The speedup presented is calculated per-GPU to better assess the effect of the data-ordering on each GPU.}
      \label{F:source_gpu_cvsf}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.48\linewidth}
      \includegraphics[width=\textwidth]{source_vector_width.pdf}
      \caption{The effect of vector-width on ``C''-ordered shallow-vectorized Intel OpenCL source term evaluations in the \avx/ CPU.}
      \label{F:source_vector_width}
  \end{subfigure}
  \caption{The effect of the \conp/ and \conv/ formulations, ``C'' and ``F'' data-ordering, and CPU vector-width on source term evaluation performance in \texttt{pyJac}.
	   It is noted the shallow-vectorized ``C'' ordered OpenCL cases corresponds to the vectorized data ordering described in~\cref{S:data}}
  \label{f:source_permutate}
\end{figure}

@@

\Cref{F:source_gpu_cvsf} shows the performance effect of the ``C'' versus ``F''-ordering on source term evaluation performance on both GPUs, with the speedup presented per-GPU.
In general, the ``C'' and ``F''-ordered shallow-vectorizations are nearly equivalent on both GPUs while the ``C'' and ``F''-ordered data are nearly equivalent for the \gpuold/ GPU, with the exception of the isopentanol model where the ``C''-ordered data is up to \SI{1.1}{$\times$} faster on the \gpunew/ (we note that there was high variablity for the \gpuold/ in this case). \todo{again, pending rerun on \gpuold/}
The rough equivalence between the ``C'' and ``F''-ordered codes GPUs is counter to what one might expect, as typically speaking coalesced memory access in a shallow-vectorization is easier to achieve if ``F''-ordering is used (see~\cref{S:data}).
However, the vectorized-data ordering here ensures that not only are memory accesses aligned to the vector width size and thus, likely coalesced. % but the improved data-locality, coupled with the larger cache sizes of the \gpunew/ GPU result in this modest speedup for ``C''-data ordering.

@@

\begin{figure}[htbp]
   \centering
  \begin{subfigure}[t]{0.48\linewidth}
      \includegraphics[width=\textwidth]{source_parallel_scaling.pdf}
      \caption{The strong parallel scaling efficiency (as defined in~\cref{e:strong_scaling}) of source term evaluation for Intel OpenCL\slash OpenMP on the \avx/ machine.}
      \label{F:source_parallel_scaling}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.48\linewidth}
      \includegraphics[width=\textwidth]{source_crossover.pdf}
      \caption{The mean run time per-condition of OpenMP source term evaluation on different cores compared to a shallow-vectorized Intel OpenCL evaluation on \num{1} (solid lines) and \num{4} (dashed lines) cores on \avx/ machine.}
      \label{F:source_crossover}
  \end{subfigure}
  \\
  \begin{subfigure}[t]{0.48\linewidth}
      \includegraphics[width=\textwidth]{source_simd_efficiency.pdf}
      \caption{The SIMD efficiency (\cref{e:simd_efficiency}) of source term evaluation for the Intel OpenCL runtime on a single core of the \avx/ CPU.}
      \label{F:source_simd_scaling}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.48\linewidth}
      \includegraphics[width=\textwidth]{source_sse_simd_efficiency.pdf}
      \caption{The SIMD efficiency (\cref{e:simd_efficiency}) of source term evaluation for the Intel OpenCL runtime on a single core of the \sse/ CPU.}
      \label{F:source_sse_simd_scaling}
  \end{subfigure}
  \caption{The parallel scaling efficiency and SIMD efficiency of source term evaluation for Intel OpenCL on the \avx/ and \sse/ CPUs.}
  \label{F:source_scaling}
\end{figure}

Finally,~\cref{F:source_scaling} displays the (strong) parallel scaling efficiency and SIMD efficiency for the CPU platforms.
The strong parallel scaling efficiency is defined as:
\begin{equation}
 \label{e:strong_scaling}
 \varepsilon = \frac{\bar{t}_{1}}{N \bar{t}_{N}}
\end{equation}
where $\bar{t}_{N}$ is the mean run time per condition on $N$ CPU cores, and $\bar{t}_{1}$ the same on a single CPU core.
The strong parallel scaling efficiency measures the speedup due to the use of additional CPU cores, as a fraction of linear speedup; strong scaling tends to decrease with the number of processors used due to memory bandwidth limitations and decreasing computation work allocated per CPU core~\cite{strong_scaling}.

\Cref{F:source_parallel_scaling} shows the strong parallel scaling efficiency of source-rate in \texttt{pyJac} on the \avx/ machine for both the shallow-vectorized Intel OpenCL and OpenMP codes.
In general, the \ce{H2}\slash\ce{CO} mechanism has the worst scaling efficiency for Intel OpenCL; although the mechanism scalings vary somewhat for OpenMP, there is no clear trend.
It is likely the poor scaling of the \ce{H2}\slash\ce{CO} for Intel OpenCL results from both its relatively small size, and few falloff\slash chemically activated reaction (in particular, the additional expensive logarithm and exponential evaluations that accompany them); i.e., there is simply not enough work to fully utilize the SIMD intructions.
Additionally, OpenMP tends to have higher scalings then the shallow-vectorized OpenCL code, e.g., \textasciitilde\num{0.9} and \textasciitilde\num{0.75} for \num{4} and \num{8} CPU cores respectively, compared to just \numrange{0.66}{0.72} and \numrange{0.44}{0.48} for OpenCL; although not pictured (to keep the figure readable), the unvectorized Intel OpenCL code has only slightly worse scaling than the OpenMP code, hence the poorer scaling is unique to the shallow-vectorizated code.
This is due to better performance of the shallow-vectorized OpenCL code, coupled with the relatively small amounts of work associated with source-rate.
To illustrate this,~\cref{F:source_crossover} shows the mean run time per-condition of the OpenMP source term evaluations for \numrange{2}{16} cores, compared to the shallow-vectorized OpenCL code on \num{1} (solid lines) and \num{4} (dashed lines) four cores on the \avx/ machine.
For all chemical models, the mean run time per-condition (and hence the computational work allocated per-core, one of the key-drivers of parallel scaling efficiency~\cite{strong_scaling}) of OpenMP running on \num{4} cores is roughly equal to that of the shallow-vectorized OpenCL code on a single core.
Similarly, OpenMP running on \num{16} cores is roughly equivalent to the OpenCL code on \num{4} cores.
Therefore, a more fair comparison of parallel scaling efficiencies is to compare OpenCL running on \num{4} cores, to OpenMP on \num{16}; the OpenMP code's parallel efficiency drops to \textasciitilde\num{0.64} for \num{16} cores, similar to OpenCL's parallel scaling efficiency of \numrange{0.66}{0.72} at \num{4} cores.
Indeed, as will be investigated in~\cref{S:jacobian_results} the strong parallel scaling efficiency for sparse Jacobian evaluation---the most computationally intensive task in this work---are far more similar between Intel OpenCL and OpenMP.
